---
title: "Homework 7"
author: "Srinidhi Raman (sr57453)"
date: "2025-04-02"
output:
   html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

Link to Github Repo: (link)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

#calling the necessary packages and libraries for this assignment
library(tidyverse)
library(ggplot2)
library(MatchIt)
```

## Problem 1: Armfolding

```{r echo=FALSE, results = "hide"}
#Load the dataset
armfolding <-  read.csv('armfold.csv')

#Count the instances of male and female
count(armfolding, Sex)

#Count the instances of male and female folding arms on top
armfolding %>% 
  count(Sex, LonR_fold)

#Computing a 95% confidence interval for the difference in proportions
confint <- prop.test(c(50, 47), c(106, 111), correct=FALSE)

```

A. There are 111 females and 106 males in this dataset. Of the 97 students who folded their left arm on top of their right arm, 47 are females, and 50 are males. The sample proportions for these two groups are 0.4234 for females and 0.4717 for males.

B. The observed difference in proportions between the two groups is 0.0483.

C. The 95% confidence interval computed for the difference in proportions is (-0.0839, 0.1805)

i. The formula for the standard error for the difference in proportions is SE(p1 - p2) = √[p1(1-p1)/n1 + p2(1-p2)/n2]

ii. The values plugged in were 0.4717 for p1, 0.4234 for p2, 106 for n1 and 111 for n2.

iii. The z* value used for this 1.96, which corresponds to the critical value for a 95% confidence interval.

D. If we were to  sample many groups of students from the same university and calculate the difference in proportions of males and females each time, we can expect that 95% of these would contain the true difference between males and females who fold their left arm first.

E. The standard error is a measure of the uncertainty in these sample populations. It measures the variability from sample to sample that we take. It measures the expected variation, with a smaller standard error resulting in a more precise observation, whereas a larger standard error will indicate that the observation may have randomly occurred.

F. In this context, the sampling distribution refers to the differences in proportions of males and females when we take repeated samples from the university, which is the population. The sample proportion number will always fluctuate from sample to sample, as will the difference in the two. However, the sample size and the true proportion that is captured within the interval will always stay the same.

G. The Central Limit Theorem is what justifies using a normal distribution to approximate the sampling distribution of the difference in sample proportions. The CLT says that if the sample is sufficiently large enough, then the distribution will be approximately normal. If the sample sizes of males and females are large enough, then the sampling distribution of the difference between their proportions will eventually be normally distributed.

H. Based on the confidence interval, I would partially agree with the person's claim. Because zero is contained within the interval, there is a possibility that there truly is no difference between males and females when it comes to arm folding. This could mean that our observed difference could truly just be random. However we cannot fully rule out the possibility of difference.

I. Yes, the confidence interval would be different across samples, because the proportion of males and females who cross their left arm over first will be bound to change every single time they survey themselves. With a larger sample size, there may not be as much of a variation, but there is a guarantee for variation every single time. If we calculated a 95% confidence interval for each of these universities, then about 95% of those intervals would have the true proportion difference in males and females.

## Problem 2: Get out the vote

```{r echo=FALSE}
#Loading the dataset
turnout <- read.csv('turnout.csv')

#Calculating the proportions of those who voted based on whether or not they got a vote
turnout %>%
  group_by(GOTV_call) %>%
  summarise(voted_yes = mean(voted1998))

#Calculating a 95% confidence interval for the difference in these two proportions
prop.test(c(sum(turnout$voted1998[turnout$GOTV_call == 1]), sum(turnout$voted1998[turnout$GOTV_call == 0])),
          c(sum(turnout$GOTV_call == 1), sum(turnout$GOTV_call == 0)))

```

### Part A

Among those who received a GOTV call, approximately 64.78% voted in 1998, and among those who did not receive a GOTV call, approximately 44.42% voted in 1998. The results of the 95% confidence interval for the difference in voting proportions between call recipients and non-recipients suggest that the true difference falls between 0.141 and 0.266.

### Part B 

To test confounding in the variables Age, MAJORPTY and voted1996, we shall look at some figures below. To test age, we will examine and analyze a histogram. For MAJORPTY, we will look at a chi-squared test of proportions, and a bar chart for votes.

```{r echo=FALSE}
#Histogram for Age
ggplot(turnout, aes(x=AGE, fill=factor(GOTV_call))) + geom_histogram(binwidth = 5, position = 'identity') + scale_fill_manual(values = c('darkorchid4','turquoise2')) + labs(title='Age Distribution by GOTV Call Status', x = 'Age', y = 'Count', fill = 'GOTV Call')

```

The histogram above shows the distribution of ages of people who did or did not receive a GOTV call. The graph shows that there is a clear majority of people who did not receive a call. Of the people who did receive a call, though, there seems to be a higher count in ages 60-80, and many of the people who did not get a call fall in the age range of 25-50. This histogram provides visual evidence that age may be a confounding factor in receiving calls and voting.

```{r echo=FALSE}
#T-test for age
t.test(AGE ~ GOTV_call, data = turnout)

```
Upon conducting a t-test for the true difference in means, we can see the confidence interval tells us that the true mean difference is somewhere in between 6.37 and 11.4 years. The p-value is 2.817e-11, which is an extremely small value, telling us that we have to reject the null. This means that age is a strong confounder, and older citizens are more likely to vote.

```{r echo=FALSE}
#bar plot for voting in 1996 vs. gotv call
ggplot(turnout, aes(x=factor(voted1996), fill=factor(GOTV_call))) + geom_bar(position='fill') + labs(title = "Proportion of Voters who got a GOTV Call vs didn't", x = "Voted in 1996", y = "Proportion", fill = "GOTV Call")

```

The bar chart above shows the proportion of votes in 1996 that recieved a call vs. didn't, and whether or not they voted in the election. The graph shows us that a higher proportion of people who did vote receeived a call compared to those who didn't. This suggests that past voting behavior could have influenced whether or not someone got a GOTV call. People who voted in 1996 were more likely to get a call, making this a confounding variable.

```{r echo=FALSE}
#Conducting a 2 sample test for the difference in proportions
prop.test(c(sum(turnout$voted1996[turnout$GOTV_call == 1]), sum(turnout$voted1996[turnout$GOTV_call == 0])),
          c(sum(turnout$GOTV_call == 1), sum(turnout$GOTV_call == 0)))

```

The two-sample test for proportions above shows us that a higher proportion of past voters received calls compared to non-past voters, which means that prior voters are more likely to receive a call than those who did not. The small p-value leads us to reject the null, and we can conclude that the variable voted1996 is a confounder.

```{r echo=FALSE}
#Chi-squared test for the majorpty variable
majorpty_table <- table(turnout$MAJORPTY, turnout$GOTV_call)
chisq.test(majorpty_table)

```

The results from the chi-squared test above tell us that the p-value is very close to 0.05, indicating that there may be very slight statistical significance. It is definitely much, much less than the other two variables. There is some evidence, though, that could be used to argue that MAJORPTY is another confounding variable.

```{r echo=FALSE}
#Confidence interval for the proportion differences
prop.test(c(sum(turnout$MAJORPTY[turnout$GOTV_call == 1]), sum(turnout$MAJORPTY[turnout$GOTV_call == 0])),
          c(sum(turnout$GOTV_call == 1), sum(turnout$GOTV_call == 0)))

```

The results of the test above show us that there is some sort of statistical significance to conclude that the variable MAJORPTY could be a confounder for whether or not people got a call to vote. However, this is a very borderline p-value, which is why we cannot strongly suggest so. But it still leans in the direction that it could be confounding.

The conclusion that we can draw from all three of these analyses is that the GOTV impact is not causal and could have been impacted by things such as age, previous voting experience and party affiliations.


### Part C

```{r echo=FALSE}
#Peforming a match with five control cases
matched_data <- matchit(GOTV_call~voted1996 + AGE + MAJORPTY, data = turnout, ratio = 5)
turnout_matched <- match.data(matched_data)

#Check for balancing
turnout_matched %>% 
  group_by(GOTV_call) %>% 
  summarise(mean_age = mean(AGE), voted_1996 = mean(voted1996), major_pty = mean(MAJORPTY))

#Check balancing for age
t.test(AGE ~ GOTV_call, data = turnout_matched)

#Check balancing for voted1996
prop.test(c(sum(turnout_matched$voted1996[turnout_matched$GOTV_call == 1]),
            sum(turnout_matched$voted1996[turnout_matched$GOTV_call == 0])),
          c(sum(turnout_matched$GOTV_call == 1), 
            sum(turnout_matched$GOTV_call == 0)))

#Check balacing for MAJORPTY
prop.test(c(sum(turnout_matched$MAJORPTY[turnout_matched$GOTV_call == 1]),
            sum(turnout_matched$MAJORPTY[turnout_matched$GOTV_call == 0])),
          c(sum(turnout_matched$GOTV_call == 1), 
            sum(turnout_matched$GOTV_call == 0)))

```

The table of summary statistics above, followed by the results of the t-test and proportion tests help us confirm that the matching was successful. The table shows us that all the means are almost identical, and all of the intervals include 0, which means that the confounding element has been removed and the matched dataset worked successfully.

```{r echo=FALSE}
#Repeating the analysis from part A with the matched data

turnout_matched %>%
  group_by(GOTV_call) %>%
  summarise(voted_yes = mean(voted1998))

# Calculating a 95% confidence interval for the difference in proportions using matched data
prop.test(c(sum(turnout_matched$voted1998[turnout_matched$GOTV_call == 1]), 
            sum(turnout_matched$voted1998[turnout_matched$GOTV_call == 0])),
          c(sum(turnout_matched$GOTV_call == 1), sum(turnout_matched$GOTV_call == 0)))

```

The results of repeating Part A with a matched dataset show us that among those who received a GOTV call, approximately 64.78% voted in 1998, and among those who did not receive a GOTV call, approximately 59.62% voted in 1998. The results of the 95% confidence interval for the difference in voting proportions between call recipients and non-recipients suggest that the true difference falls between 0.010 and 0.147.

Overall, we can conclude that the Get Out The Vote call had a positive effect on voting in the 1998 election based on the results from the matched data. The reduction of inflation caused by the confounding variables shows a more accurate measure of the call's impact.